Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-01-27 docs/manuscript/manuscript.tex
Reorganization of contribution section.

Lack of water accessibility is a major crisis for many vulnerable people around the world. We have determined to seek literature reviews from the past 15 years to have the most up to date information because information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility in a global scale. Therefore we can conduct a research to fill many major gaps  to analyze the water accessibility predictor variables which could increase and enhance water accessibility in global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically. 

@@

Improving water accessibility have been a priority by many organizations around the world, realizing it is the basic human rights to have clean and accessible drinkable water in past 15 years. During 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fullerTrackingProgressGlobal2016}. However, some researchers raises concerns on lack of clarity with the global water accessibility monitoring \citep{ondaGlobalAccessSafe2012, jiaHighResolutionSpatialDistribution2016}. However, in an era of rapidly changing advanced technologies, international, and government policies, some of the older research might not be adequate for up to date information for  example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest SDGs of providing clean and accessible drinking water for all by 2030, researchers have devoted their study to understand water accessibility in global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,wssccWaterSanitationHygiene2014,  ondaGlobalAccessSafe2012, bartramGlobalMonitoringWater2014, ondaCountryClusteringApplied2014, whoDrinkingwater,fullerTrackingProgressGlobal2016, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there has been studies focusing on local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagalsDoesImprovedAccess2006, geereDomesticWaterCarrying2010, simaWaterFlowsEnergy2013, hoChallengeGlobalWater2014, sakisakaIntroductionCommunityWater2015, jiaHighResolutionSpatialDistribution2016}. Where majority raised concerns about lack of monitoring sustainable water accessibility measurements for many people, and even few argues for greater focus in many people's access to drinking water sources. Nonetheless, there is a lack of robust study on modes of transportation related to enhance water accessibility for all people by 2030. 

@@

\citet{cassiviAccessDrinkingWater2018} highlights about how there should be more information to the effect of including a 30-minute
collection time to monitor access to drinking water. The lack of access to on-premises water sources can cause the use
of greater distant alternative sources. Therefore causing the quantity of retrieved water to reduce as the water plateau
phenomenon.\citet{cassiviAccessDrinkingWater2018} hypothesizes that the households who must travel further are using unimproved
sources. For our research, we can discover what modes of transportation people are traveling to retrieve the water and
in what quantity depending on their means of transportation. Therefore, learning how to decrease the travel time of
people to retrieve the water because according to the article, it can be a great burden and does not increase the
quality of one's life to retrieve water daily and far away from their homes.

\citet{tortajadaAchievingUniversalAccess2018} highlights that sharing the immense improvement at the 2015 convention negatively
affected many people because they actually have not acquired clean drinking water. While 2.6 billion people gained
access to drinking water, they did not have constructive feedback on the quality of the water. Also, our research can
fill in the blank on how did these people actually retrieve the water? Our research can measure the performance of how
people retrieve water which may not be in their premises.

\citet{amitMeasuringAffordabilityAccess2019} highlights their data analysis similar to what we will be doing and found out that for
non-piped households, collection costs are 22\% of the coping costs, while collection costs for piped households are
less than 2\% of the coping costs. The researchers also realize that there is economic inequality because the majority
of the piped households are in households that are considered wealthy households and vice versa. Our research can
pinpoint how households are traveling to collect water to improve potential policy-making for water accessibility.

\citet{priceDifferenceDayCan2019} highlights the temporal dynamics of drinking water access and quality in urban slums which
is home to about 1 billion people globally. \citet{priceDifferenceDayCan2019} discusses the temporal changes in drinking water
access and quality in urban slums may influence on health risks. \citet{priceDifferenceDayCan2019} believes that monitoring
temporal dynamics should be considered over three interlinked time scales: short, medium and long term.\citet{priceDifferenceDayCan2019} recommends future research to explore the short-term water access and quality monitoring. Also we should learn to recognize the role of socio-cultural factors that may impact the temporal dynamics of safe water access. Our project can pinpoint additional water accessibility variables that can be used to focus to recognize the time variations as well.

\citet{rawasComparingUtilityreportedHours2020} highlights the need for different approaches of intermittent water supplies (IWS) to monitor piped water supply perpetually in Peru. This relates and informs one mode of transportation of clean and accessible water. Our research can make inferences about if the customers receive clean water elsewhere, such as water tankers, kiosks, vendors, etc. Overall, the article found that the percent of households with (IWS) was a few percentage points greater than of the reported utilities.

\citet{deshpandeMappingGeographicalInequalities2020} discusses their findings of improved overall water availability and piped water accessibility for the globe (40\% to 50.3\%). \citet{deshpandeMappingGeographicalInequalities2020} also found that it was the lowest increase in sub-Saharan Africa, where the accessibility was mostly concentrated in urban regions. \citet{deshpandeMappingGeographicalInequalities2020} focuses mostly on understanding trends in diarrhea burden, where the needs are greatest for the improvement of access to safe drinking water and sanitation facilities. Our study can benefit this research by implementing the best and effective modes of transportation of clean water to improve safe drinking water and sanitation. Even with the water pipe access increasing, they can't conclude if the people had clean water as they needed right away. Overall, the access to safe drinking water and sanitation improved globally between 2000 and 2017, but dis-proportionality undermines reaching the Sustainable Development Goals.

\citet{GoalCleanWater} explains specific targets that have goals and indicators to track global water accessibility. The article quantifies and provides Graphs: safe drinking water, safe sanitation, and hygiene, safe sanitation and hygiene, ambient water quality, water use efficiency, levels of freshwater stress, integrated water management, trans-boundary water cooperation, protect and restore water-related ecosystems, water and sanitation support, local participation in sanitation management. They have many ambitious plans to achieve by 2020 and 2030, and our research can find a correlation to why achieving this may be possible.

\citet{cassiviEvaluatingSelfreportedMeasures2021} focuses a case study in Southern Malawi, where data collected in March and April of 2019 at the conclusion of the rainy season. The sample was consisted of 375 households without access to water on their premises.   The article highlighted using Euclidean Distance measurement is more reliable than the self-reported time. Also queuing time measurement can be hard to be measured because of their ambiguity. Their research gives greater insight into how do we define water accessibility and how increase of water accessibility can impact people's lives beneficially.

% citep (for citations fully enclosed in parentheses)
% cite (for citations without any parentheses)
% citet (for citations where the year only is in parentheses)
%%\textit{Discuss the who/what/when/where? (You already have this... just need to reorganize or improve structure)}
%Another example is \citet{fuller2016tracking}.
%\citet{rawas2020comparing} highlights the need for different approaches of intermittent water supplies (IWS) to monitor piped water supply perpetually in Peru.
%\citet{cassivi2018access} discuss time access...
%\hl{Perhaps insert a table summarizing the various dimensions of water accessibility that have been studied}
%Other examples: \citep{amit2019measuring,ho2014challenge}
We are interested in analyzing the impact of transportation modes (usage/ownership) on water accessibility. We also desire to see an implementation of this statistical research piloted in one of the countries to see if the water accessibility variables predictors can improve the water accessibility for its country.

>>>>

Lack of water accessibility is a major crisis for many vulnerable people around the world. We have determined to seek literature reviews from the past 15 years to have the most up to date information because information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility on a global scale. Therefore we can conduct research to fill many major gaps to analyze the water accessibility predictor variables which could increase and enhance water accessibility on a global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically. 

@@

Improving water accessibility has been a priority by many organizations around the world, realizing it is the basic human right to have clean and accessible drinkable water in the past 15 years. From 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fullerTrackingProgressGlobal2016}. However, some researchers raise concerns on lack of clarity with the global water accessibility monitoring \citep{ondaGlobalAccessSafe2012, jiaHighResolutionSpatialDistribution2016}. However, in an era of rapidly changing advanced technologies, international, and government policies, some of the older research might not be adequate for up to date information for example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest SDGs of providing clean and accessible drinking water for all by 2030, researchers have devoted their studies to understand water accessibility at global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,wssccWaterSanitationHygiene2014,  ondaGlobalAccessSafe2012, bartramGlobalMonitoringWater2014, ondaCountryClusteringApplied2014, whoDrinkingwater,fullerTrackingProgressGlobal2016, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there have been studies focusing on the local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagalsDoesImprovedAccess2006, geereDomesticWaterCarrying2010, simaWaterFlowsEnergy2013, hoChallengeGlobalWater2014, sakisakaIntroductionCommunityWater2015, jiaHighResolutionSpatialDistribution2016}. Where majority raised concerns about the lack of monitoring sustainable water accessibility measurements for many people and even a few argue for greater focus on many people's access to drinking water sources. Nonetheless, there is a lack of robust study on modes of transportation-related to enhance water accessibility for all people by 2030. 

@@

However, in the most recent years, there is a lack of information that needs to focus on people to receive accessible drinking water. The relation between water accessibility and the total time to retrieve drinking water is a concern and (SDGs) should emphasize this study \citep{cassiviAccessDrinkingWater2018}. In the most recent years, there needs a greater understanding of the water accessibility indicators to enhance drinking water accessibility on a global scale for many vulnerable people. While research on water accessibility on a global scale has diminished due to an increase of water for many people \citep{fullerTrackingProgressGlobal2016}, but \citet{deshpandeMappingGeographicalInequalities2020} raises concerns that poorer community around the world are still not receiving clean and accessible drinking water. Several research and studies focus on a global scale to understand water accessibility and sanitation \citep{cassiviAccessDrinkingWater2018, tortajadaAchievingUniversalAccess2018, priceDifferenceDayCan2019,deshpandeMappingGeographicalInequalities2020}. Some research highlighted the coping cost in small community in India \citep{amitMeasuringAffordabilityAccess2019} and researched on inequality with water accessibility and sanitation \citet{amitMeasuringAffordabilityAccess2019,priceDifferenceDayCan2019, rawasComparingUtilityreportedHours2020}. As many research journal articles focused on different aspects of water accessibility and sanitation in particular countries also brought greater insight and necessary to understand water accessibility variables \citep{rawasComparingUtilityreportedHours2020, cassiviEvaluatingSelfreportedMeasures2021}. \citet{GoalCleanWater} emphasized providing safe and accessible drinking water for all on a global scale, which defines that we need a greater understanding of water accessibility and what water accessibility variables be investigated further. Therefore, we will be conducting robust research to understand water accessibility variables and predictors to improve water accessibility to bring clean and safe drinking water for all by 2030 as (SDGs) visioned.

The journal articles were missing what modes of transportation may enhance water accessibility. Therefore we will be analyzing the impact of transportation modes (usage/ownership) on water accessibility. We will therefore research by statistical analysis in a global scale and pilot in one of the countries to determine if the water accessibility variables predictors can improve the water accessibility for its country.

% citep (for citations fully enclosed in parentheses)
% cite (for citations without any parentheses)
% citet (for citations where the year only is in parentheses)
%\textit{Discuss the who/what/when/where? (You already have this... just need to reorganize or improve structure)}
